NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-01-26 manuscriptPGPE.tex
Update MD methods and results and discussion.

\begin{abstract}
% insert abstract here
  Chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles. Because lipid membranes are in liquid state under physiological conditions, the conformational ensembles of lipid headgroups and their dependence on chemistry are difficult to resolve experimentally. Here, we combine solid state NMR experiments and molecular dynamics simulations from NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of key lipid types in liquid lamellar phase under various biologically relevant conditions. Interpretation of NMR experiments using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in neutral and charged cellular membranes. However, the populations of different conformations are dictated by the headgroup chemistry. Together with the analysis of protein-bound lipids from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations independently on the headgroup chemistry. Therefore, the selective adsorption of proteins to membranes is likely regulated by specific protein--lipid interactions, rather than conformational restrictions of lipids. Our results pave the way to comprehensive understanding of lipid--protein interaction energetics, and the understanding of complex biomolecular assemblies such as membrane proteins.

@@

Chemical compositions of hydrophilic lipid headgroups vary between different
organelles and organisms, and different lipid types
regulate protein functions in many different ways \cite{lee03,vanmeer08}.
Lipids can directly bind to proteins or indirectly affect protein
functions by altering membrane properties or charge \cite{lee03,lemmon08}.
While the specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{lee11,lemmon08},
it is not clear if the specificity is driven by the differences in accessible
conformational states between lipid types or
by specific intermolecular lipid--protein interactions.

The conformational ensembles of lipids in the physiologically relevant lamellar liquid phase
are typically derived from NMR experiments, particularly from 
C--H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}. Notably, such measurements can be performed also on living cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone conformations are largely similar irrespectively of the headgroup \cite{gally81}, 
and the headgroup conformations are similar in PC, PE and PG lipids, while the headgroup is more rigid in PS lipids \cite{wohlgemuth80,buldt81}. 
However, these tendencies are based only on the absolute values, whereas
the necessity of order parameter signs in capturing the conformational ensembles of lipids has been recently demonstrated \cite{botan15,ollila16,ferreira16}.
Furthermore, the detailed understanding of lipid conformational ensembles
is limited by the lack of universal models that would map order parameters to structural ensembles \cite{pezeshkian18,akutsu20}.

@@

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the studied lipids specifically bind to various proteins \cite{yeagle14}.
Furthermore, we use our results to elucidate lipid--protein interactions and the effect of charges on lipid conformations.

@@

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of POPE and POPG
were measured using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings
using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The NMR experiments were identical as in our previous work~\cite{antila19}.
POPE (1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine) and POPG (1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol)) powder were purchased from Avanti polar lipids. The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

@@

Molecular dynamics simulation data were collected using
the NMRlipids Open Collaboration \cite{botan15}, which is running at
\url{nmrlipids.blogspot.fi} and \url{github.com/NMRlipids/NMRlipidsIVotherHGs}.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables~\ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Tables~\ref{systemsMIX} and \ref{systemsMIX2}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

\subsection{Analysis of molecular dynamics simulation data}
The large set of MD simulation data was analysed using the development version of NMRlipids databank.
Unique naming convention for lipid atoms in each force field was defined using the mapping files
and analysis for all simulations indexed in NMRlipids databank manner were performed
using Python codes.

The C--H bond order parameters were calculated directly
from the carbon and hydrogen positions using the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\left\langle 3\cos^2\theta -1 \right\rangle,
\end{equation}
where $\theta$ is the angle between the C--H bond and the membrane normal
(taken to align with $z$, with bilayer periodicity in the $xy$-plane).
Angular brackets denote average over all sampled configurations.
The order parameters were first calculated averaging over time separately
for each lipid in the system. The average and
the standard error of the mean were then calculated over different lipids.
Python programs that use the MDAnalysis library \cite{agrawal11,gowers16}
used for all atom simulations are available in Ref.~\citenum{MATCHgit}
(\texttt{scripts/calcOrderParameters.py}). For united atom simulations, the trajectories
with hydrogens having ideal geometry were constructed first using either \texttt{buildH} program~\cite{buildH}
or (\texttt{scratch/opAAUA\_prod.py}) in  Ref.~\citenum{MATCHgit}, and the order parameters were
then calculated from these trajectories. This approach has been tested against trajectories
with explicit hydrogens and the deviations in order parameters are small \cite{buildH,piggot17}.\\
%\todo{BuildH program is now cited with a direct link to the GitHub repo. I think that a release to Zenodo would be nice in the final publication.}\\
The number density profiles of ions were calculated using the \texttt{gmx density} tool
of the GROMACS sofware package \cite{gromacsMANUAL}.

@@

\subsection{Analysis of conformations of protein-bound lipids}
Lipid structures from Protein Data Bank (PDB, \url{http://www.rcsb.org/})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api}).
First, all PDB entries containing PC, PE, PG, or PS lipid headgroups were collected.
The ligand names to identify the lipids were:
PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW for PC;
8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE for PE;
PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT for PG; and
PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP for PS.
Secondly, all PDB structures containing these ligands were downloaded and the dihedral angles of the first lipid in the latest version of the structures within each PDB code were calculated using the MDTraj python library \cite{mcgibbon15}.
The used Jupyter notebook is available from the project's GitHub repository (\textit{ scripts/pdbSEARCH.ipynb}).

@@

\begin{figure*}[bt]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from CHARMM36 simulations demonstrating the differences in conformational ensembles between different lipids. 
     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     \textbf{C)} Distributions of heavy atom dihedral angles of different lipids from CHARMM36 simulations.
     \textbf{D)} Distributions of P-N vector angle with respect to membrane normal.
  }
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids
that are not bound to proteins in an electrostatically neutral cell membrane,
we measured the C--H bond order parameters and their signs from of POPG and POPE
in the liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the data in Figs.~\ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C--H bonds, except for the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig.~\ref{HGorderParameters}B)).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig.~\ref{HGorderParameters}B)).
This interpretation is confirmed by SIMPSON calculations in Fig.~\ref{POPGsimpson}.

@@

To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C--H bond order parameters from different MD force fields to the experimental data. None of the force fields reproduced the order parameters within experimental accuracy for any of the lipids analysed in this work (Figs.~\ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}. Nevertheless, the CHARMM36 force field performs best for all the headgroups and reproduces the experimentally observed distinct order parameters of PS and PG lipids (Fig.~\ref{structures}B). To understand the structural origin of these distinct order parameters, we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in Fig.~\ref{structures}C). Major differences between headgroups are observed for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of PS lipids for which distinct distributions are observed particularly around phosphate group. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids, which could also explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

In conclusion, all lipid headgroups sample very broad conformational ensembles in liquid lamellar phase.
Between headgroups, the sampled dihedral angles are within approximately same ranges for all headgroup types. Therefore, we suggest that all lipid headgroups are very flexible---thereby being
able to adopt wide range of multiple conformations when interacting with proteins, ions or other biomolecules. Our results support the models that propose free rotation around phosphate group for PC lipids, which decouples the dynamics and conformations between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group. The wide range of observed conformations suggest that the structures in lipid crystals \cite{buldt81,pascher92} play only a minor role, and that the models aiming to explain NMR data using only few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in liquid lamellar state.

@@

To resolve lipid headgroup conformational ensembles in cell membrane bearing positive charge, we first analyzed CHARMM36 simulations which correctly capture the experimentally measured decrease in PC headgroup order parameters upon addition of cationic surfactants into a bilayer in figure \ref{changesWITHsurf} A).
Heavy atom dihedral angle distributions in figure \ref{changesWITHsurf} B
reveal that the decrease of trans state probabilities in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$
dihedrals are the major effects of positive charge on lipid headgroup conformations.
Choline region remains essentially unchanged and only minor changes are observed in other dihedrals
even though almost half of the molecules in membranes are cationic surfactants.
We also tried to resolve PC lipid headgroup conformational ensembles in mixed membrane with
negatively charged lipids \cite{scherer87},
but the accuracy of currently available force fields
was not sufficient for this (Fig. \ref{HGorderparametersPCvsPEPG}).

Also binding of ions may affect the lipid headgroup conformational ensembles in physiological conditions.
In line with the results from cationic surfactants, 
the decreased probability for the trans state of g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral is the main structural consequence of
bound ion to PC headgroup (Fig.~\ref{DIHSwithCAlipid17eccPOPC} and \ref{DIHSwithCAcharmm36POPC}) in
Lipid17ecc (having the most realistic response of PC lipid headgroup order parameters to the calcium binding in Fig.~\ref{changesWITHCaClPG})
and CHARMM36 simulations.
%simulations with the ECC included, where
%Similar but smaller effect is seen in CHARMM36 simulations with less bound calcium (Fig. \ref{DIHSwithCAcharmm36POPC}).
%Also the headgroup order parameters of
Response of charged lipids to the ion binding is less well understood and more difficult to capture in MD simulations \cite{antila19,melcr20}.
The most realistic simulations (Lipid17 and Slipids in Fig.~\ref{changesWITHCaClPG}) suggest
%istic response of PG $\beta$-carbon to CaCl$_2$  even thought the calcium binding affinity was overestimated in these models.
%These models suggest
that the upward tilting of the headgroup dipole upon addition of CaCl$_2$ is weaker in PG than in PC headgroup,
%(\ref{changesWITHCaClPG}),
but dihedral distributions are more sensitive to the bound ions in PG (Figs.~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
However, the changes in dihedrals upon addition of calcium differ between the best models (Figs.~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}),
none of the simulations captures the calcium binding affinity and conformational ensemble of PG lipids simultaneously, and experimental data to evaluate the response of $\alpha$-carbon order parameters to the added calcium in PG is not available.
Therefore, more accurate force fields are required for the detailed analysis of the interactions between ions and charged lipids, as concluded previously also for PS lipids \cite{antila19,melcr20}.

Despite the limited capability of simulations to interpret some of the experimental data due to
the inaccuries in force fields, we conclude that the experimentally observed changes
in headgroup order parameters upon addition of charges into bilayer arise from relatively
small changes in conformational ensembles. These changes eventuate from
mild changes in dihedral angle distributions, rather than from restriction of lipids into
fixed conformations. Therefore, lipid headgroups remain in disordered state sampling large
space of different conformations, thereby being able to interact with different molecules
in various ways, also in membranes bearing electric charge.

@@

To analyze lipid conformations in protein-bound states,
we calculated the heavy atom dihedral distributions from 
all lipid conformations available within protein structures
deposited in the PDB \cite{berman00} in Fig.~\ref{dihedralsFROMpdb}.
We found 176 PC, 198 PE, 70 PG, and 41 PS conformations, which
present lipids that are tightly bound to proteins in fixed conformations
that can be determined as a part of protein structure using crystallography, NMR, or cryo-EM.

>>>>

\begin{abstract}
% insert abstract here
  Chemistry of lipid headgroups, the water facing components of cell membranes that regulate cell functions via lipid--protein interactions, varies between organisms and organelles. Because lipid membranes are in liquid state under physiological conditions, the conformational ensembles of different lipid headgroups are difficult to resolve experimentally. Here, we combine solid state NMR experiments and molecular dynamics simulations from NMRlipids open collaboration to resolve the conformational ensembles of the headgroups of key lipid types in liquid lamellar phase under various biologically relevant conditions. Interpretation of NMR experiments using the plethora of simulation data collected in the NMRlipids project suggests that all lipid headgroups sample a wide range of conformations in neutral and charged cellular membranes. However, the populations of different conformations are dictated by the headgroup chemistry. Together with the analysis of protein-bound lipids from the protein data bank (PDB), this suggests that lipids can bind to proteins in wide range of conformations independently on the headgroup chemistry. Therefore, the selective adsorption of proteins to membranes is likely regulated by specific protein--lipid interactions, rather than conformational restrictions of lipids. Our results pave the way to comprehensive understanding of lipid--protein interaction energetics, and the understanding of complex biomolecular assemblies such as membrane proteins.

@@

Chemical compositions of hydrophilic lipid headgroups vary between different
organelles and organisms, and different lipid types
regulate protein functions in many different ways \cite{lee03,vanmeer08}.
Lipids can directly bind to proteins or indirectly affect protein
functions by altering membrane properties such as charge or elasticity \cite{lee03,lemmon08}.
While the specific interactions with certain lipid headgroups
are known to be essential for the function of several proteins \cite{lee11,lemmon08},
it is not clear if the specificity is driven by the differences in accessible
conformational states between lipid types or
by specific intermolecular lipid--protein interactions.

The conformational ensembles of lipids in the physiologically relevant lamellar liquid phase
are typically derived from NMR experiments, particularly from 
C--H bond order parameters measured using $^2$H NMR~\cite{seelig77c,davis83,Semchyschyn04}. Notably, such measurements can be performed also on living cells \cite{gally81,scherer87,seelig90}.
These studies suggest that the glycerol backbone conformations are largely similar irrespectively of the headgroup \cite{gally81}, 
and the headgroup conformations are similar in  phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG) lipids, while the headgroup is more rigid in phosphatidylserine (PS) lipids \cite{wohlgemuth80,buldt81}. 
However, these tendencies are based only on the absolute values, whereas
the necessity of order parameter signs in capturing the conformational ensembles of lipids has been recently demonstrated \cite{botan15,ollila16,ferreira16}.
Furthermore, the detailed understanding of lipid conformational ensembles
is limited by the lack of universal models that would map order parameters to structural ensembles \cite{pezeshkian18,akutsu20}.

@@

Here, we use natural abundance $^{13}$C NMR experiments and MD simulations from the NMRlipids open collaboration
to resolve the differences in conformational ensembles between PC, PE, PG and PS lipid headgroups.
Zwitterionic PC and PE are the most common lipids in eukaryotes and bacteria, respectively \cite{vanmeer08,sohlenkamp16}.
PE is also the second most abundant glycerophospholipid in eukaryotic cells
and has been related to various diseases \cite{vance15,calzada16,patel17}.
PS and PG are the most common negatively charged lipids in eukaryotes and bacteria, respectively,
and affect membrane protein functionality and signaling \cite{lemmon08,leventis10,sohlenkamp16,hariharan18}.
All the studied lipids specifically bind to various proteins \cite{yeagle14}. We use our results to elucidate also lipid--protein interactions and the effect of charges on lipid conformations.

@@

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of 1-palmitoyl-2-oleoyl-sn-glycero-3-phosphoethanolamine (POPE) and 1-palmitoyl-2-oleoyl-sn-glycero-3-phospho-(1'-rac-glycerol) (POPG), purchased from Avanti polar lipids, were measured using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}. The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
The NMR experiments were identical as in our previous work~\cite{antila19}. The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

@@

Molecular dynamics simulation data were collected and analyzed using
the methods from the NMRlipids Open Collaboration project (\url{nmrlipids.blogspot.fi}) \cite{botan15,catte16,ollila16,antila19}.
%\url{github.com/NMRlipids/NMRlipidsIVotherHGs}. 
Simulation and accessibility details of more than 70 systems simulated for this work are given in the supplementary information. Quality evaluation of these simulations enables us to select the best models for the interpretation of lipid conformational ensembles from the experimental data.

The quality of lipid headgroup and glycerol backbone conformational ensembles in PE and PG simulations were evaluated using the C-H bond order parameters \cite{botan15}. Interactions between different headgroups in simulations of mixed bilayers were evaluated monitoring the changes in headgroup order parameters upon mixing the lipids \cite{antila19}. The ion binding affinities and response of lipids to bound charge were evaluated monitoring the changes of lipid headgroup order parameters \cite{catte16,antila19}.

@@

\subsection{Analysis of conformations of protein-bound lipids}
\todo{This is to be finished once the final analysis is done as discussed in this issue: https://github.com/NMRLipids/NMRlipidsIVPEandPG/issues/40}

Lipid structures from Protein Data Bank (PDB, \url{http://www.rcsb.org/})
were searched using PDBe REST API (\url{www.ebi.ac.uk/pdbe/pdbe-rest-api}).
First, all PDB entries containing PC, PE, PG, or PS lipid headgroups were collected.
The ligand names to identify the lipids were:
PLC, PX4, 6PL, LIO, HGX, PC7, PC8, P1O, 6O8, XP5, EGY, PLD, SBM, HXG, and PCW for PC;
8PE, PTY, 3PE, PEH, PEF, 6OE, 6O9, 9PE, PEV, 46E, SBJ, L9Q, PEK, EPH, ZPE, 9TL, 9Y0, 6OU, LOP, and PEE for PE;
PGT, PGK, LHG, 44G, PGV, OZ2, D3D, PGW, DR9, P6L, PG8, H3T, and GOT for PG; and
PSF, PS6, Q3G, P5S, D39, PS2, 17F, and 8SP for PS.
Secondly, all PDB structures containing these ligands were downloaded and the dihedral angles of the first lipid in the latest version of the structures within each PDB code were calculated using the MDTraj python library \cite{mcgibbon15}.
The used Jupyter notebook is available from the project's GitHub repository (\textit{ scripts/pdbSEARCH.ipynb}).

@@

\begin{figure*}[bt]
  \centering
   \includegraphics[width=18.0cm]{./Figs/figure2.eps}
   \caption{\label{structures}
     Results from the best simulation model (CHARMM36) simulations demonstrating the differences in conformational ensembles between different lipids. 
     \textbf{A)} Snapshots with overlayed C$_\beta$, C$_\alpha$ and O$_\alpha$ atoms and occurence of different conformations.
     \textbf{B)} Headgroup and glycerol backbone region order parameters of different different lipids.
     \textbf{C)} Distributions of heavy atom dihedral angles of different lipids from simulations.
     \textbf{D)} Distributions of P-N vector angle with respect to membrane normal.
  }
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids that are not bound to proteins in an electrostatically neutral cell membrane, we measured the C--H bond order parameters and their signs of POPG and POPE in the liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the data in Figs.~\ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C--H bonds, except for the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig.~\ref{HGorderParameters}B)).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig.~\ref{HGorderParameters}B)).
This interpretation is confirmed by SIMPSON calculations in Fig.~\ref{POPGsimpson}.

@@

%To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C--H bond order parameters from different MD force fields to the experimental data. 
%Although none of the simulations reproduce the headgroup order parameters within experimental accuracy for any of the analysed lipids (Figs.~\ref{HGorderParametersPE} and \ref{HGorderParametersPOPG} in the SI and Refs. \cite{botan15,antila19}), the CHARMM36 
%force field performs best for all the headgroups and 
%simulations reproduce the experimentally observed distinct order parameters of PS and PG lipids (Fig.~\ref{structures}B). 
To understand the structural origin of distinct order parameters for PG and PS lipids, we selected the simulations that best reproduce the differences between headgroups (based on the quality evaluation in the SI and Refs. \cite{botan15,antila19}) and calculated the distributions of heavy atom dihedral angles from simulations, shown in Fig.~\ref{structures}C. Major differences between headgroups are observed for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of PS lipids for which distinct distributions are observed particularly around phosphate group. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids. This may explain the more rigid headgroup structures in PS~\cite{browning80,buldt81}. Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

All lipid headgroups sample very broad conformational ensembles in liquid lamellar phase and sampled dihedral angles are within approximately same ranges for all headgroup types (Fig. \ref{structures}). The results suggest that all lipid headgroups are very flexible---thereby being able to adopt wide range of multiple conformations when interacting with proteins, ions or other biomolecules. Our results support the models that propose free rotation around phosphate group for PC lipids, which decouples the dynamics and conformations between acyl chains and headgroups \cite{klauda08c,antila21b}, and suggests that these models could be applied for all lipids containing the phosphorus group. The wide range of observed conformations suggest that the structures in lipid crystals \cite{buldt81,pascher92} play only a minor role, and that the models aiming to explain NMR data using only few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20} are not sufficient to capture the large conformational space of lipids in liquid lamellar state.

@@

To resolve lipid headgroup conformational ensembles in cell membrane bearing positive charge, we analyzed the simulations (CHARMM36) that correctly capture the experimentally measured decrease in PC headgroup order parameters upon addition of cationic surfactants into a bilayer in figure \ref{changesWITHsurf} A). Heavy atom dihedral angle distributions in figure \ref{changesWITHsurf} B reveal that the decrease of trans state probabilities in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$
dihedrals are the major effects of positive charge on lipid headgroup conformations. Choline region remains essentially unchanged and only minor changes are observed in other dihedrals even though almost half of the molecules in membranes are cationic surfactants.

Also binding of ions may affect the lipid headgroup conformational ensembles in physiological conditions. In line with the results from cationic surfactants, the decreased probability for the trans state of g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral is the main structural consequence of
bound Ca$^{2+}$ ion to PC headgroup (Fig.~\ref{DIHSwithCAlipid17eccPOPC} and \ref{DIHSwithCAcharmm36POPC}) in the most realistic MD simulations models (lipid17ecc and CHARMM36 in Fig.~\ref{changesWITHCaClPG}).
%Lipid17ecc (having the most realistic response of PC lipid headgroup order parameters to the calcium binding in Fig.~\ref{changesWITHCaClPG}) and CHARMM36 simulations. 
%Response of charged lipids to the ion binding is less well understood and more difficult to capture in MD simulations \cite{antila19,melcr20}.
The most realistic simulations suggest
%istic response of PG $\beta$-carbon to CaCl$_2$  even thought the calcium binding affinity was overestimated in these models.
%These models suggest
that the upward tilting of the headgroup dipole upon addition of CaCl$_2$ is weaker in PG than in PC headgroup,
%(\ref{changesWITHCaClPG}),
but dihedral distributions are more sensitive to the bound ions in PG (Lipid17 and Slipids in Figs.~\ref{changesWITHCaClPG},~\ref{DIHSwithCAslipidsPOPG} and \ref{DIHSwithCAlipid17POPG}).
However, the changes in dihedrals upon addition of calcium differ between the best models (Figs.~\ref{DIHSwithCAslipidsPOPG} and~\ref{DIHSwithCAlipid17POPG}), none of the simulations captures the calcium binding affinity and conformational ensemble of PG lipids simultaneously, and experimental data to evaluate the response of $\alpha$-carbon order parameters to the added calcium in PG is not available. Also the lipid headgroup conformational ensembles in mixed PC:PE and PC:PG membranes were not possible to resolve with the currently available force fields and experimental data (Figs.~\ref{HGorderparametersPCvsPEPG} and~\ref{HGorderparametersPGvsPCchange}).

%Therefore, more accurate force fields are required for the detailed analysis of the interactions between ions and charged lipids, as concluded previously also for PS lipids \cite{antila19,melcr20}.

Despite the 
%limited capability of simulations to interpret some of the experimental data due to the 
inaccuries in simulations in some cases, we can conclude that the experimentally observed changes in headgroup order parameters upon addition of charges into bilayer arise from relatively small changes in conformational ensembles. These changes eventuate from mild changes in dihedral angle distributions, rather than from restriction of lipids into
fixed conformations. Therefore, lipid headgroups remain in disordered state sampling large space of different conformations, thereby being able to interact with different molecules in various ways, also in membranes bearing electric charge.

@@

To analyze the alterations in lipid conformations when bound to proteins,
we calculated the heavy atom dihedral distributions from lipid conformations within protein structures deposited in the PDB \cite{berman00} in Fig.~\ref{dihedralsFROMpdb}.
\todo{This part is to be finished once the analysis if finalized}
We found 176 PC, 198 PE, 70 PG, and 41 PS conformations, which
present lipids that are tightly bound to proteins in fixed conformations
that can be determined as a part of protein structure using crystallography or cryo-EM.
